A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-09-01 paper/project.tex
Single rod updated data files, edits to figures and text

The simple yet complex water molecule is one of the most important 
substances for life and industry. The structure and geometry of the 
water molecule leads to a strong polarity and complex intermolecular
forces. The hydrogen bonds formed by water are very strong, creating
tetragonal clusters (CITE?) that lead to many of water's unusual properties,
such as the high melting, boiling, and critical temperatures.
The large dielectric constant (D$\sim$80) (CITE) is partially responsible for
water as a universal solvent, along with the ease of availability and use.

@@

It is challenging to describe all the relevant properties of water using
a model, especially if a low computational cost is also desired.

Some models, including SAFT, are designed for use with
homogeneous systems only. Although these models are useful, much of the interesting
chemistry that occurs in water cannot be described with a homogeneous only model. To
describe solvation and other processes in aqueous solution, a more robust theory is 
needed.

Classical density-functional theory (DFT) provides an efficient way to evaluate the
free energy using the density of the fluid. The free energy is a functional of the 
density function and is independent of any external potentials\cite{ebner1976}. 
Mermin\cite{mermin1965thermal} gives a theorem extending that of Hohenberg and 
Kohn\cite{hohenberg1964inhomogeneous} to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r) d^3r\right\}
\end{equation}
This is the basis for classical DFT, which can be used with a variety of external potentials, 
from the simple interaction with a 
hydrophobic wall to the complex strong interaction with hydrophilic solutes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory and Methods}
We introduce a new classical density functional for water that
predicts the \needsworklater{vapor density, liquid density, bulk
  modulus, and bulk surface tension, as well as allows for
  liquid-vapor coexistence}.  The Helmholtz free energy functional is
based on SAFT, and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n]  +
F_\textit{disp}[n]+ F_\textit{assoc}[n],
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
a hard-sphere free energy, $F_\textit{assoc}$ is the free energy of
association, and $F_\textit{disp}$ is the dispersion energy.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional,
which incorporates effects of kinetic energy.  The ideal gas
functional is given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{n(\xx)} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules.  The ideal gas free
energy functional has the property that the contact density at a hard
surface $n_c$ is proportional to the pressure on that surface,
according to the contact value theorem
\begin{equation}\label{contactvaluethm}
  p = n_c k_BT \:,
\end{equation}
which also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes.  These properties are retained by the total functional,
provided the remaining terms are \emph{purely
  nonlocal}~\cite{ashcroft?}.

@@

\subsection{Association free energy}
The first attractive energy term is the association term, which
accounts for hydrogen bonding.  Hydrogen bonds are modeled as four
attractive patches (``association sites'') on the surface of the hard
sphere.  These four sites represent two protons and two electron lone
pairs.  There is an attractive energy $\epsilon_\textit{assoc}$ when
two molecules are arranged such that the proton of one overlaps
with the lone pair of the other.  The volume over which this
interaction occurs is $\kappa_\textit{assoc}$, giving the association
term in the free energy has two empirical parameters that are fit to
experimental data.

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/saturated-liquid-density}
% \end{center}
% \caption{The theoretical versus experimental saturated liquid density
%   versus temperature.  }
% \label{fig:saturated-liquid-density}
% \end{figure}

\subsection{Dispersion free energy}
The final term in the free energy is the dispersion term.  We use a
dispersion term based on the SAFT-VR
approach\cite{gil-villegas-1997-SAFT-VR}, which has two free
parameters, an interaction energy $\epsilon_\textit{disp}$ and a
length scale $\lambda_\textit{disp}$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
\end{center}
\caption{The theoretical versus experimental temperature versus density. }
\label{fig:temperature-vs-density}
\end{figure}

% \begin{figure}
% \begin{center}
% \includegraphics[width=\columnwidth]{figs/entropy}
% \end{center}
% \caption{The free energy, internal energy, and temperature*entropy 
% versus density at room temperature.  }
% \label{fig:energy-room-temp}
% \end{figure}

@@

\fixme{I'm not sure if this paragraph belongs here.  Somewhere I think
  it's worth discussing any of these homogeneous plots we include, and
  this might be the right place.  We also might want to include one or
  more of them earlier in the text.  A general rule is that any figure
  that is included should be referenced in the text.  If there's
  nowhere in the text that it fits, then it probably doesn't belong in
  the paper....  So I'm not removing the follwoing bits since it's up
  to you, Jess, whether to remove the figures.  We should
  \emph{definitely} remove some of them.}  For reference, we plot the
predicted vapor pressure (see Figure~\ref{fig:equation-of-state}) and
coexistence curve (see
Figure~\ref{fig:temperature-vs-density}) against experimental
values.  To find the densities, we use the ``common tangent'' rule,
as illustrated in Figure~\ref{fig:homogeneous}.

% \begin{table}
% \begin{tabular}{ccc}
%   Fitting Parameters& \hspace{2em} & Experimental Values\\
% \begin{tabular}{cc}
%   $R$ & 2.3 $a_o$ \\
%   $A$ & -0.0100909 Hartree \\
%   $\gamma$ & 5\\
%   $V$ & 515.792 $a_o^3$  \\
%   $b$ & 0.75 $a_o$ \\
%   $\mu$ & -0.014063 Hartree 
% \end{tabular}
% && 
% \begin{tabular}{cc}
%  $n_l$ & 4.939 $\times 10^{-3} a_o^{-3}$\\
%  $n_v$ & 1.141 $\times 10^{-7} a_o^{-3}$\\
%  $\sigma$ & ??? \\
%  $B$ & $7.435\times 10^{-5}$ Hartree$/a_o^3$\\
%  $T$ & 298.15 K \\
% \end{tabular}
% \end{tabular}
% 
% \caption{Empirical parameters of optimized functional and experimental
%   values matched, where $a_o$ is the Bohr radius, and $\sigma$ is the
%   surface tension.  This table is not yet complete, but it should
%   be soon...  We'll also need to think about the best units to use
%   here.  Atom units might be best, but it also might be better to
%   consistently use eV and angstrom or something.  I'm not sure about
%   putting $k_BT$ data in the table.  Maybe it belongs in the caption?}
% \label{tab:parameters}
% \end{table}

@@

\subsection{One hydrophobic rod}

We move into two dimensions by studying a single hydrophobic rod
immersed in water. Figure \ref{fig:density-single-rod} shows density
profiles for three different radii, 0.3 nm, 0.5 nm, and 1 nm. There are the
expected oscillations outside the rod and as the distance away from the rod increases
the density approaches that of the bulk saturated liquid density. The small radius
rod has comparatively larger oscillations and a higher contact density.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Energy versus diameter for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in Figure 
\ref{fig:surface-tension}. (FIX - 0.4 nm?) }
\label{fig:energy-vs-diameter}
\end{figure}

\subsubsection{Resolution and convergence tests}

Figure \ref{fig:densityresolution} shows the density profiles for three different
values of the lattice spacing resolution. The density oscillations around a single
rod of radius 0.5 nm look nearly identical for all resolutions except for the first peak. 
This contact density is important as it will be proportional to the pressure on the 
rod\ref{contactvaluethm}. The 
contact density calculated using the medium resolution is within less than 4\% of the value at
high resolution, while the low resolution contact density is off by about 5\%. There is some
uncertainty in the value of contact density due to the different lattice spacings. The total
free energy calculations compare much more favorably. The medium resolution is within 
0.5\% and the low resolution is within 1\% of the high resolution free energy.

@@

@@

Here we will want to cite~\cite{walther2004hydrodynamic}, which gives
simulations for two interacting nanotubes in water.  It's not
precisely the same, but if we can't find something more similar, this
would be worth citing.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two identical hydrophobic rods with diameters
of 1.0 nm. }
\label{fig:rods-energy-vs-distance}
\end{figure}

>>>>

The simple yet complex water molecule is one of the most important 
substances for life and industry. The structure and geometry of the 
water molecule leads to a strong polarity and complex intermolecular
forces. The hydrogen bonds formed by water are very strong, creating
tetragonal clusters that lead to many of water's unusual properties,
such as the high melting, boiling, and critical temperatures as well as the
density maximum above the melting point.
The large dielectric constant is partially responsible for
water acting as a universal solvent, along with the ease of availability and
use.
There are many models used to describe water since so many different
processes occur in aqueous solution. It is challenging to describe all the 
relevant properties of water at once, especially if a low computational cost is
also desired.

Some models are primarily designed for use with homogeneous systems. Although
these 
models are useful in predicting some bulk properties, the interesting
chemistry that occurs in water cannot be described with a homogeneous only
model. To
describe solvation and other processes in aqueous solution, a more robust theory
that 
can handle varying densities is needed.

@@

Classical density-functional theory (DFT) provides an efficient way to evaluate
the free energy using the density of the fluid. The free energy is a functional
of the density function and is independent of any external
potentials\cite{ebner1976}. The most common use of DFT is calculating ground
state electron densities in quantum mechanical systems. The Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} uses DFT to write the ground state
energy in terms of a universal functional of the electron density and a separate
term including the external potential. Mermin\cite{mermin1965thermal} gives a
theorem extending that of Hohenberg and Kohn to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r)
d^3r\right\}
\end{equation}
This is the basis for classical DFT, where the relevant density is not the
electron density,
but the density of molecules. This can be used with a variety of 
external potentials, from the simple interaction with a 
hydrophobic wall to the more complex, stronger interaction with hydrophilic
solutes.

\subsection{Statistical Associating Fluid Theory}

The high hydrogen bonding and strong polar interactions in water characterize it
as an associating fluid, and thus it requires a different treatment than simple
fluids which are mainly defined by van der Waals forces and weak electrostatic
interations. The Statistical Associating Fluid Theory (SAFT) was designed for
use with fluids such as water, to account for the important association forces
(like hydrogen bonds) as well as any effects due to clusters or chains forming
in the liquid\cite{muller2001molecular}. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory and Methods}
We introduce a new classical density functional for water that
predicts the vapor density, liquid density, and bulk surface tension, as well 
as allows for liquid-vapor coexistence.  The Helmholtz free energy functional is
based on SAFT, and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n]  +
F_\textit{disp}[n]+ F_\textit{assoc}[n],
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
a hard-sphere free energy, $F_\textit{disp}$ is the dispersion energy,
and $F_\textit{assoc}$ is the free energy of association.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional,
which incorporates effects of kinetic energy.  The ideal gas
functional is given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{n(\xx)} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules.  The ideal gas free
energy functional has the property that the contact density at a hard
surface $n_c$ is proportional to the pressure on that surface,
according to the contact value theorem
\begin{equation}\label{contactvaluethm}
  p = n_c k_BT \:,
\end{equation}
which also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes.  These properties are retained by the total functional,
provided the remaining terms are \emph{purely
  nonlocal}~\cite{ashcroft?}\needsworknow{I can't find this citation anywhere!}.

@@

\subsection{Dispersion free energy}
The first term in the attractive energy is the dispersion term.  We use a
dispersion term based on the SAFT-VR
approach\cite{gil-villegas-1997-SAFT-VR}, which has two free
parameters, an interaction energy $\epsilon_\textit{disp}$ and a
length scale $\lambda_\textit{disp}$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
\end{center}
\caption{The theoretical versus experimental coexistence curve. Away from
the critical temperature, the agreement is good. }
\label{fig:temperature-vs-density}
\end{figure}

\subsection{Association free energy}
The final attractive energy term is the association term, which
accounts for hydrogen bonding.  Hydrogen bonds are modeled as four
attractive patches (``association sites'') on the surface of the hard
sphere.  These four sites represent two protons and two electron lone
pairs.  There is an attractive energy $\epsilon_\textit{assoc}$ when
two molecules are arranged such that the proton of one overlaps
with the lone pair of the other.  The volume over which this
interaction occurs is $\kappa_\textit{assoc}$, giving the association
term in the free energy has two empirical parameters that are fit to
experimental data.

@@

For reference, we plot the predicted vapor pressure (see
Figure~\ref{fig:equation-of-state}) and coexistence curve (see
Figure~\ref{fig:temperature-vs-density}) against experimental values.  These
match those calculated by Clark \emph{et al}. To find the densities, we use the
``common tangent'' rule, as illustrated in Figure~\ref{fig:homogeneous}.
\needsworknow{Expand the previous paragraph}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/energy-vs-diameter}
\end{center}
\caption{ Energy versus diameter for a single hydrophobic rod
immersed in water. This should have an asymptote equal to the surface
tension at room temperature, and it agrees well with the surface tension in
Figure \ref{fig:surface-tension}. }
\label{fig:energy-vs-diameter}
\end{figure}

@@

\subsection{One hydrophobic rod}
We move into two dimensions by studying a single hydrophobic rod
immersed in water. Figure \ref{fig:density-single-rod} shows density
profiles for three different radii, 0.3 nm, 0.5 nm, and 1 nm. There are the
expected oscillations outside the rod and as the distance away from the rod
increases
the density approaches that of the bulk saturated liquid density. The small
radius
rod has comparatively larger oscillations and a higher contact density.

Figure \ref{fig:energy-vs-diameter} shows the free energy per area versus
radius for the range of rods studied, with radii from 0.05 nm to 1.5 nm. After 
about 0.5 nm or so, this reaches
an asymptote equal to the surface tension. From the inset in Figure 
\ref{fig:surface-tension}, it is clear that the surface tension at room 
temperature is approximately 72 mN/m. This is very close to the value from
the larger diameter single rod energy per area, which is about 75~mN/m. 
\needsworknow{Here it will be beneficial to add some discussion of any
relevant experimental work or other simulations that deal with one rod-like 
thing in water. Do we have anything currently matching this?}

\subsubsection{Resolution and convergence tests}
Figure \ref{fig:densityresolution} shows the density profiles for three
different values of the lattice spacing resolution. The density oscillations
around a single rod of radius 0.5 nm look nearly identical for all resolutions
except for the first peak. This contact density is important as it will be
proportional to the pressure on the rod (see equation \ref{contactvaluethm}). The contact
density calculated using the medium resolution is within less than 4\% of the
value at high resolution, while the low resolution contact density is off by
about 5\%. There is some uncertainty in the value of contact density due to the
different lattice spacings. The total free energy calculations compare much more
favorably. The medium resolution is within 0.5\% and the low resolution is
within 1\% of the high resolution free energy.

@@

Here we will want to cite~\cite{walther2004hydrodynamic}, which gives simulations for 
two interacting
nanotubes in water.  It's not precisely the same, but if we can't find something
more similar, this would be worth citing - they use CNTS with diameter 1.25 nm
and find drying occurs for separations less than 0.9-1nm.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/rods-energy-vs-distance}
\end{center}
\caption{ Energy versus separation for two hydrophobic rods. Radii 0.3nm, 0.5nm, 1.0nm
 }
\label{fig:rods-energy-vs-distance}
\end{figure}
